Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-07 agda.tex
final popl paper + supplement

This development is available in the supplemental material. At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete. We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward. The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{talk about determinism either here or there}

@@

>>>>

This development is available in the supplemental material. At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete. We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward. The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.

@@

\subsubsection{Determinism}
A useful property would be \emph{action determinism}, i.e. that performing an action produces a unique result. Formally, this would be established as follows:
% \begin{theorem}[Action Determinism] All of the following hold:
% \label{thrm:actdet}
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

  % \begin{quote}
  %   \begin{enumerate}
  %   \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
  %   \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
  %   \item $\tcompat{\htau}{\htau'}$, and
  %   \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
  %     $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
  %   \end{enumerate}
  % \end{quote}
  % hold, then $\zexp' = \zexp''$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
  \zexp''$.
\end{enumerate}
% \end{theorem}

This is not a theorem of the system as described in this paper. The reason is quite subtle: the semantics of two construction actions, $\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the analytic case than they do in the synthetic case -- the difference has to do with what the generated ascription contains. Both cases can technically ``fire'' due to action subsumption. The mechanization includes a ``proof'' of determinism for all cases but these two. Resolving this ambiguity formally requires some rather inelegant machinery to exclude these particular base cases from action subsumption, so we are content to admonish implementors who apply action subsumption when another analytic rule would do.

### 2016-07-07 formally.tex
final popl paper + supplement

Our overview below omits certain ``uninteresting'' details. The supplement includes the complete collection of rules, in definitional order. These rules, along with the proofs of the metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement. We will summarize this effort in Sec. \ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.

@@

Types and expressions that contain no holes are \emph{complete types} and \emph{complete expressions}, respectively. Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec. \ref{sec:future} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers. %Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.

@@

The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that appears as a premise in the subsumption rule is a reflexive and symmetric (but not transitive) relation between H-types defined judgementally in Figure \ref{fig:type-consistency}. This relation coincides with equality for complete H-types. Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type. The type hole is therefore compatible with every type. This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e. unknown) type \cite{Siek06a}.\todo{fwd reference related work}

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions, $\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this form, so lambda abstractions can appear only in analytic position, i.e. where an expected type is known.\footnote{It is possible to also define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.} Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible solutions to this problem. One solution would be to define a second rule specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}. This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement\todo{is admissible the right word}. Encouragingly, this judgement also arises in the study of gradual type systems\todo{first paper?}.

@@

Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application. It also makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole. Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}. Subsequent work on gradual typing introduced this notion of type matching\todo{cite}.

@@

The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on. All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found. Any other sub-term is ``dotted'', i.e. it is either an H-type or H-expression. Taken together, every Z-type and Z-expression contains exactly one focused H-type or H-expression by construction. This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)

@@

The rules for relative movement within Z-expressions are similarly straightforward. Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib in supplement too}
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
% \begin{figure}
% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
% \begin{displaymath}
% \begin{array}{@{}rcl}
%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
%   %\\
%   \TABperformMove
%       {\zwsel{\hexp : \htau}}
%       {\aMove{\dChild}}
%       {      \zwsel{\hexp} : \htau}
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%     {
%       \hexp : \zwsel{\htau}
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dNext}
%     }{
%       \hexp : \zwsel{\htau}
%     }
%   \\[2mm]
%   \TABperformMove
%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hinj{i}{\hexp}}}
%       {\aMove{\dChild}}
%       {\hinj{i}{\zwsel{\hexp}}}      
%   \\
%   \TABperformMove
%       {\hinj{i}{\zwsel{\hexp}}}      
%       {\aMove{\dParent}}
%       {\zwsel{\hinj{i}{\hexp}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dChild}}
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
% \end{array}
% \end{displaymath}
% \caption{Movement action semantics for sum type forms.}
% \end{figure}
The expression movement judgement is defined as follows.

@@

This requires developing machinery for reasoning about sequences of actions. There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements. To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.\todo{put stars on these?}

A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.\todo{define this in supplement}

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
\begin{equation}\label{rule:performAna-construct-lam-2}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:\todo{supp}
\begin{equation}
  \label{r:coneapfn}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hexp}{\zwsel{\hehole}}}
      {\htau_2}
  }
\end{equation}
If the expression under the cursor synthesizes a type that is inconsistent with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tehole}
  }
\end{equation}

@@

\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty hole with the number expression $\hnum{n}$. If the empty hole is being asked to synthesize a type, then the rule is straightforward:
\begin{equation}
  \label{r:conenumnum}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\fnumlit{n}}}
      {\zwsel{\hnum{n}}}
      {\tnum}
  }
\end{equation}
If the empty hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside the hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\fnumlit{n}}}
      {\hhole{\zwsel{\hnum{n}}}}
  }
\end{equation}

@@

\todo{make sure there is no more compatibility or focus words}If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against. This amounts to analyzing the enveloped expression against the provided type (by subsumption):
\begin{equation}\label{r:finishana}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

@@

\begin{subequations}
In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
\begin{equation}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\htau}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

@@

When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
\begin{equation}\label{rule:zipper-asc}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}

>>>>

Our overview below omits certain ``uninteresting'' details. The supplement includes the complete collection of rules, in definitional order. These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement. We will summarize this effort in Sec. \ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.

@@

Types and expressions that contain no holes are \emph{complete types} and \emph{complete expressions}, respectively. Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec. \ref{sec:rw} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers. %Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.

@@

The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that appears as a premise in the subsumption rule is a reflexive and symmetric (but not transitive) relation between H-types defined judgementally in Figure \ref{fig:type-consistency}. This relation coincides with equality for complete H-types. Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type. The type hole is therefore consistent with every type. This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e. unknown) type \cite{Siek06a}.

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions, $\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this form, so lambda abstractions can appear only in analytic position, i.e. where an expected type is known.\footnote{It is possible to also define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.} Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible solutions to this problem. One solution would be to define a second rule specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}. This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly, this judgement also arises in the study of gradual type systems \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

@@

Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application. It also makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole. Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}. Later work on gradual typing introduced this notion of type matching to resolve this redundancy.

@@

The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on. All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found. Any other sub-term is ``dotted'', i.e. it is either an H-type or H-expression. Taken together, every Z-type and Z-expression contains exactly one selected H-type or H-expression by construction. This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)

@@

The rules for relative movement within Z-expressions are similarly straightforward. Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
% \begin{figure}
% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
% \begin{displaymath}
% \begin{array}{@{}rcl}
%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
%   %\\
%   \TABperformMove
%       {\zwsel{\hexp : \htau}}
%       {\aMove{\dChild}}
%       {      \zwsel{\hexp} : \htau}
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%     {
%       \hexp : \zwsel{\htau}
%     }{
%       \aMove{\dParent}
%     }{
%       \zwsel{\hexp : \htau}
%     }
%   \\
%   \TABperformMove
%       {
%       \zwsel{\hexp} : \htau
%     }{
%       \aMove{\dNext}
%     }{
%       \hexp : \zwsel{\htau}
%     }
%   \\[2mm]
%   \TABperformMove
%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hinj{i}{\hexp}}}
%       {\aMove{\dChild}}
%       {\hinj{i}{\zwsel{\hexp}}}      
%   \\
%   \TABperformMove
%       {\hinj{i}{\zwsel{\hexp}}}      
%       {\aMove{\dParent}}
%       {\zwsel{\hinj{i}{\hexp}}}
%   \\[2mm]
%   \TABperformMove
%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dChild}}
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dNext}}
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dPrev}}
%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
%   \\
%   \TABperformMove
%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
%   \\
%   \TABperformMove
%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
%       {\aMove{\dParent}}
%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
% \end{array}
% \end{displaymath}
% \caption{Movement action semantics for sum type forms.}
% \end{figure}
The expression movement judgement is defined as follows.

@@

This requires developing machinery for reasoning about sequences of actions. There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements. To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.

A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
is inserted inside a hole, which defers the type inconsistency as previously discussed:
\begin{equation}\label{rule:performAna-construct-lam-2}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:
\begin{equation}
  \label{r:coneapfn}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hexp}{\zwsel{\hehole}}}
      {\htau_2}
  }
\end{equation}
If the expression under the cursor synthesizes a type that is inconsistent with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tehole}
  }
\end{equation}

@@

\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty hole with the number expression $\hnum{n}$. If the empty hole is being asked to synthesize a type, then the rule is straightforward:
\begin{equation}
  \label{r:conenumnum}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\fnumlit{n}}}
      {\zwsel{\hnum{n}}}
      {\tnum}
  }
\end{equation}
If the empty hole is being analyzed against a type that is inconsistent with $\tnum$, then we must place the number expression inside the hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\fnumlit{n}}}
      {\hhole{\zwsel{\hnum{n}}}}
  }
\end{equation}

@@

If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against. This amounts to analyzing the enveloped expression against the provided type (by subsumption):
\begin{equation}\label{r:finishana}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

@@

\begin{subequations}
In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):
\begin{equation}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\htau}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

@@

When the cursor is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
\begin{equation}\label{rule:zipper-asc}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}
